---
title: "Differential analysis with HTSet"
author: Chenghao Zhu
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
table.table {
    white-space: nowrap;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(edgeR)
library(limma)
library(DESeq2)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(HTSet)
```

## introduction

High through-put experiments such as RNA-seq, proteomics, and metaboliomics often generate hundrands to millions of features. For experiments such as metabolomics and proteomics in MRM mode that generates continous data, linear model is usually perferred, while for experiments that generates counts data, such as RNA-seq, the negative binomial generalized linear model methods are popular. In this vigeneete, we will go through several popular methods and models that are widely used in fields such as RNA-seq. The function [`model_fit()`](https://zhuchcn.github.io/HTSet/reference/model_fit.html) is a wrapper of three bioconductor packages [limma](https://bioconductor.org/packages/release/bioc/html/limma.html), [edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html), and [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) that performs differnetial analysis on a `HTSet` object and returns the result in an organized structure.

The basic usage of [`model_fit()`](https://zhuchcn.github.io/HTSet/reference/model_fit.html) is below.

```{r, eval=FALSE}
model_fit(object, design, coef, engine)
```

`object` must be a `HTSet` object. `design` is the study design matrix, usually returned by the `model.matrix` function. `coef` is the coefficient to perform statistical test. `engine` must be one of the three backend statistical package mentioned above. 

The HTSet package has two dataset included. They both can be loaded using the `data()` function. 

```{r}
data(exrna)
data(lipidome)
```

In this vignette, we will go though some examples of using `model_fit()` to perform statistical analysis with the two datasets.

## Backend engines

### count data

High through-put experimental data generated by nucleiic acid sequencing are were usually presented as counts per gene (RNA-seq) or counts per ASV/OTU (16S-seq). Counts data by nature do not fit gaussian distribution, thus linear model is not directly applicable. Thus, the major approches are (1) transforming counts data to continous data and normalized it, and (2) use negative binomial generalized linear model (glm). For the former approch, counts data are often transformed into logCPM followed by linear modeling. Another or a better approach is using [limma](https://bioconductor.org/packages/release/bioc/html/limma.html)'s `voom` function, which estimates the mean-variance relationship of the log-counts, and generates a precision weight for each observation. As for negative binomial glm, [edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html) and [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) two widely recognized packages.

### exrna

The dataset [`exrna`](https://zhuchcn.github.io/HTSet/reference/exrna.html) is the small RNA-seq result of isolated HDL from human plasma of 3 patients with systemic lupus erythematosus and 3 healthy control. The raw sequence fastq files were downloaded from GEO ([GSE121865](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE121865)), and processed using the [exceRNApipeline](https://github.com/zhuchcn/exceRNApipeline), using STAR agaist hg38. Gene counts were generated using [htseq-count](https://github.com/simon-anders/htseq) agaist gencode's annotation. 

```{r}
exrna
```

```{r}
kable(exrna$pdata[,c(1,2,4,7)], format = "html") %>%
    kable_styling()%>%
    column_spec(5, background = "orange") %>%
    scroll_box(width = "100%")
```

### limma voom

We first create a study design matrix using the `model.matrix` function.

```{r}
# create the design matrix
design = model.matrix( ~ Condition, data = exrna$pdata)
kable(design)
```

We are going to test whether each gene is expressed differently between patinets and control, thus the cofficient that we are going to test should be `ConditionSystemic Lupus Erythematosus` (second column name). We can than call `model_fit`. Setting `voom = TRUE` tells `model_fit` to run `voom` before calling limma's `lmFit` function.

```{r}
coef = "ConditionSystemic Lupus Erythematosus"
fit1 = model_fit(exrna, design = design, coef = coef, engine = "limma", voom = TRUE)
```

```{r}
fit1$results[order(fit1$results$pval),] %>% head %>% kable()
```

Draw a volcano plot

```{r}
volcanoplot(fit1, hline = 0.05)
```

P value histogram

```{r}
hist(fit1, vline = 0.05)
```

### limma (but not voom)

It is also possible to transform count data into logCPM, and use limma directly.

```{r}
exrna_cpm = exrna
exrna_cpm$edata = cpm(exrna_cpm$edata, prior.count = 1, log = TRUE)
fit2 = model_fit(exrna_cpm, design = design, coef = coef, engine = "limma")
```

```{r}
fit2$results[order(fit2$results$pval),] %>% head %>% kable()
```

Drawing a volcano plot

```{r}
volcanoplot(fit2, hline = 0.05)
```

P value histogram

```{r}
hist(fit2, vline = 0.05)
```

### DESeq2

To use DESeq2 to perform negative binomial glm, simply set `engine = "DESeq2"`

```{r}
fit3 = model_fit(exrna, design = design, coef = coef, engine = "DESeq2")
```

```{r}
fit3$results[order(fit3$results$pval),] %>% head %>% kable()
```

Drawing a volcano plot

```{r}
volcanoplot(fit3, hline = 0.05)
```

P value histogram

```{r}
hist(fit3, vline = 0.05)
```

### edgeR

Similar to DESeq2, edgeR also performes negative binomial glm test. Two models are provided by edgeR, the quasi-likelihood test and likelihood ratio test.

Quasi-likelihood test

Using the wrapper function `model_fit` in HTSet

```{r}
fit4 = model_fit(exrna, design = design, coef = coef, engine = "edgeR", edger_model = "qlf")
fit4$results[order(fit4$results$pval),] %>% head %>% kable()
```

Drawing a volcano plot

```{r}
volcanoplot(fit4, hline = 0.05)
```

P value histogram

```{r}
hist(fit4, vline = 0.05)
```

Likelihood ratio test

```{r}
fit5 = model_fit(exrna, design = design, coef = coef, engine = "edgeR", edger_model = "lrt")
fit5$results[order(fit5$results$pval),] %>% head %>% kable()
```

Drawing a volcano plot

```{r}
volcanoplot(fit5, hline = 0.05)
```

P value histogram

```{r}
hist(fit5, vline = 0.05)
```

## Complex models

### lipidome

`model_fit()` also supports more complex study design models. The dataset [`lipidome`](https://zhuchcn.github.io/HTSet/reference/lipidome.html) is the result of a lipidomics analysis on isolated HDL from 10 human subjects in a 2x2 randomized cross-over study. The study was published on *Metabolomics*, 2019 (doi: [10.1007/s11306-019-1579-1](doi.org/10.1007/s11306-019-1579-1)). 

```{r}
lipidome
```

```{r}
summary(lipidome$pdata[,c("Treatment", "Timepoint")])
```

In the examples above, we were simply comparing disease patients to control. However in this study, we are asking a slightly more complex question. We would like to see how each lipid species change before and after the two treatments FF and Med. And since each subject was sampled four times, the model that we are using would be as blew. And $\beta_3$ is the coefficient that we would like to test.

$$y = \beta_0 + \beta_1 Treatment + \beta_2 Timepoint + \beta_3 Timepoint*Treatment + \beta_4 Subject + e$$

The design matrix should be as below.

```{r}
# in R's equation, ~ Treatment * Timepoint is a shorthand for 
# ~ Treatment + Timepoint + Treatment * Timepoint
design = model.matrix(~ Treatment * Timepoint + Subject + 1, data = lipidome$pdata)
coef = "TreatmentMed:TimepointPre"
```

Because the `lipidome` is continous data, the limma can be directly applied.

```{r}
lipidome2 = transform_by_sample(lipidome, function(x) log(x/sum(x)))
fit6 = model_fit(lipidome2, design = design, coef = coef, engine = "limma")
fit6$results[order(fit6$results$pval),] %>% head %>% kable
```

```{r}
volcanoplot(fit6, hline = 0.05)
```

```{r}
hist(fit6, vline = 0.05)
```

## The End

`model_fit()` is simply a wrapper for the statistical pipelines from limma, edgeR, and DESeq2 with a clean interface and output format. You can certainly use limma, edgeR or DESeq2 directly and it is also not too complicated. Sometimes different models may come up with slightly different gene/feature list, however, as long as the model is approciate for the type of data to answer the specific question, your conclusion should be affected much. On thing to bear in mind that although most high through-put experiments are exploratory and hypothesis generating, running all different kinds of models and keep the one with the gene list that you like the most is probably a bad practice and may lead to inappropriate conclusion.
